---
title: "eTape Calibration Investigation"
author: "A.J. Brown"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
# load libraries
library(rethinking)
library(dplyr)
library(tidyr)
library(tidyverse)
```

## Import and Clean Data
```{r}
setwd("C:/Users/ansle/OneDrive/Documents/GitHub/etape-calibration-investigation")
#d <- read.csv("data/etape_data.csv")
d <- read.csv("data/etape_data_plus_etape_study.csv")

# calculate water depth in centimeters
d <- d %>%
  mutate(water_depth_cm = water_depth_inch * 2.54)

# remove "bad" data from "good/bad" column
d <- d %>%
  filter(good.bad == "good") %>%
  select(-good.bad)

# categorize columns as the appropriate type


#Column Name	Description
#year	Year of the test
#water_depth_inch	Known water depth during calibration (in inches)
#resistivity_ohm	Measured resistance from eTape sensor (in ohms)
#etape_ID	Unique identifier for each physical eTape sensor
#etape_length	Nominal length of the eTape sensor (e.g., 15 for 15-inch sensor)
#good/bad	Quality flag for measurement ('good' or 'bad')
#notes	Optional notes regarding conditions or sensor status
d <- d %>%
  mutate(
    year = as.factor(year),
    water_depth_inch = as.numeric(water_depth_inch),
    water_depth_cm = as.numeric(water_depth_cm),
    resistivity_ohm = as.numeric(resistivity_ohm),
    etape_ID = as.factor(etape_ID),
    etape_length = as.factor(etape_length),
  )

# prepare data for analysis in rethinking by converting to a list
d_real <- list(
  Y = d$year,
  W = d$water_depth_cm,
  R = d$resistivity_ohm,
  L = d$etape_length,
  I = d$etape_ID,
  # list items with lengths
  K_Y = length(unique(d$year)),
  K_W = length(unique(d$water_depth_cm)),
  K_R = length(unique(d$resistivity_ohm)),
  K_L = length(unique(d$etape_length)),
  K_I = length(unique(d$etape_ID)),
  # Prior for alpha (intercept) for each etape length
  aPrior = c(28, 32, 40, 48, 61)  
)

# Print Summary Statistics for each etape length
summary_stats <- d %>%
  group_by(etape_length, water_depth_inch) %>%
  summarise(
    mean_resistivity = mean(resistivity_ohm, na.rm = TRUE),
    median_resistivity = median(resistivity_ohm, na.rm = TRUE),
    sd_resistivity = sd(resistivity_ohm, na.rm = TRUE),
    min_resistivity = min(resistivity_ohm, na.rm = TRUE),
    max_resistivity = max(resistivity_ohm, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  ) %>%
  arrange(etape_length, water_depth_inch)

# Output summary statistics
print(summary_stats)
```

## Simulated Dataset Generation
```{r}
set.seed(123)

# 1. Define true parameter values for each eTape length
etape_lengths   <- c(8, 12, 15, 18, 24)
true_intercepts <- c("8" = 28, '12' = 37, '15' = 46, '18' = 52, '24' = 70)   # ohms at 0 in
true_slopes     <- c("8" = -0.010, '12' = -0.014, '15' = -0.017, '18' = -0.02, '24' = -0.025)  # ohms per cm
true_sds        <- c("8" = 1, '12' = 1, '15' = 0.5, '18' = 2, '24' = 1.5)  # error SD

# 2. Simulate data for each depth and etape length
n_per_depth <- 100
depths_inch <- 1:24

sim_data <- expand.grid(
  etape_length     = etape_lengths,
  water_depth_inch = depths_inch,
  replicate        = 1:n_per_depth,
  KEEP.OUT.ATTRS   = FALSE
)

# 3. Calculate mean resistivity for each row using its etape_length and depth
sim_data <- sim_data %>%
  mutate(
    water_depth_cm  = water_depth_inch * 2.54,
    etape_length_chr = as.character(etape_length),
    r_mu = (water_depth_cm - true_intercepts[etape_length_chr]) / true_slopes[etape_length_chr],
    resistivity_ohm = rnorm(n(), mean = r_mu, sd = true_sds[etape_length_chr])
  ) %>%
  select(
    etape_length,
    water_depth_inch,
    water_depth_cm,
    resistivity_ohm
  )

# 4. Summarize just like real data (by etape_length and depth)
summary_stats_sim <- sim_data %>%
  group_by(etape_length, water_depth_inch) %>%
  summarise(
    mean_resistivity   = mean(resistivity_ohm, na.rm = TRUE),
    median_resistivity = median(resistivity_ohm, na.rm = TRUE),
    sd_resistivity     = sd(resistivity_ohm, na.rm = TRUE),
    min_resistivity    = min(resistivity_ohm, na.rm = TRUE),
    max_resistivity    = max(resistivity_ohm, na.rm = TRUE),
    count              = n(),
    .groups = "drop"
  ) %>%
  arrange(etape_length, water_depth_inch)

# 5. Print or view the simulated summary table
print(head(summary_stats_sim, 20))  # show first 20 rows for preview

# 6. Convert etape_length to a factor and create index
sim_data <- sim_data %>%
  mutate(
    etape_length_f = factor(etape_length, levels = etape_lengths),  # levels in correct order
    L = as.integer(etape_length_f)
  )

# Save the mapping for reference if needed
etape_levels <- levels(sim_data$etape_length_f)  # c("12", "15", "18", "24")

# Prepare the list for the ulam model
d_sim <- list(
  W   = sim_data$water_depth_cm,   # Response: depth in cm
  R   = sim_data$resistivity_ohm,  # Predictor: resistivity
  L   = as.factor(sim_data$L),     # Integer index (1:4) for etape length
  K_L = length(etape_lengths),    # Number of unique etape lengths (should be 4)
  aPrior = c(28, 32, 40, 48, 61)  # Prior for alpha (intercept) for each etape length
)

```


## Model Generation and Calibration
```{r}
# Fit Model with Simulated Data
m1.0 <- ulam(
  alist(
    # Observation model
    W ~ dnorm(mu, sigma_L),
    
    # Mean structure: interaction between analyte and sampler type
    mu <- alpha[L] + beta[L]*R,
    
    # Priors for eTape model-specific parameters
    alpha[L] ~ normal(aPrior, 1),  # Intercept for each etape length
    beta[L] ~ normal(-0.17, 1),  # Slope for each etape length
    
    # Prior for measurement error
    sigma_L ~ exponential(1)

  ),
  data = d_sim,
  chains = 4,
  cores = 4
)
```


### Model Calibration Results - Sim Data
```{r}
p_sim <- precis(m1.0, depth = 2)
p_sim
plot(p_sim)
```

```{r}
traceplot(m1.0)
```

## Re-calibrate model using real data
```{r}
real_model <- update(m1.0, data = d_real)
```

### Model Calibration Results - Real Data
```{r}
p <- precis(real_model, depth = 2, digits = 5)
p
plot(p)
```

```{r}
trankplot(real_model)
traceplot(real_model)
```

## Results 
```{r}
# Extract posterior samples
#post <- extract.samples(m1.0)
post <- extract.samples(real_model)

# Use the rethinking color palette
colset <- c("#8080FF", "#F98400", "#00A08A", "#E2AD00", "#800080", "#008000")
etape_labels <- c("8 in", "12 in", "15 in", "18 in", "24 in")
K_L <- length(etape_labels)
```

Alpha
```{r}
# Posterior densities for alpha (Intercept) by eTape length
# Set up empty plot area
plot(NULL, xlim = range(post$alpha), ylim = c(0, 2),
     xlab = expression(alpha ~ "(Intercept, cm)"),
     ylab = "Density",
     main = "Posterior Density: alpha by eTape length")

# Overlay densities for each eTape length using dens()
for (k in 1:K_L) {
  dens(post$alpha[, k], col = colset[k], lwd = 3, add = TRUE)
}
legend("topright", legend = etape_labels, col = colset, lwd = 3, bty = "n")

```

Beta
```{r}
# Posterior densities for beta (Slope) by eTape length
# Set up empty plot area
plot(NULL, xlim = range(post$beta), ylim = c(0, 3000),
     xlab = expression(beta ~ "(Slope, cm/立)"),
     ylab = "Density",
     main = "Posterior Density: beta by eTape length")

# Overlay densities for each eTape length using dens()
for (k in 1:K_L) {
  dens(post$beta[, k], col = colset[k], lwd = 3, add = TRUE)
}
legend("topright", legend = etape_labels, col = colset, lwd = 3, bty = "n")


```

Generative model predictions
```{r}
# Function to plot predictions for each eTape length at a specified resistivity
plot_depth_predictions <- function(R_new, post, etape_lengths, colset, ...) {
  n_etape <- length(etape_lengths)
  n_draws <- nrow(post$alpha)
  preds <- matrix(NA, nrow = n_draws, ncol = n_etape)
  
  for (i in 1:n_etape) {
    preds[, i] <- post$alpha[, i] + post$beta[, i] * R_new
  }
  
  # Plotting
  all_preds <- as.vector(preds)
  x_min <- min(all_preds)
  x_max <- max(all_preds)
  percent_buffer <- 0.1
  xlim <- c(x_min - percent_buffer * abs(x_min), x_max + percent_buffer * abs(x_max))
  y_max <- max(apply(preds, 2, function(col) max(density(col)$y)))
  ylim <- c(0, y_max * 1.1)
  plot(NULL, xlim = xlim, ylim = ylim,
       xlab = "Predicted Water Depth (cm)", ylab = "Density",
       main = sprintf("Posterior Water Depth | Resistivity = %s 立", R_new), ...)
  for (i in 1:n_etape) {
    dens(preds[, i], col = colset[i], lwd = 3, add = TRUE)
  }
  legend("topright", legend = paste(etape_lengths, "in"), col = colset, lwd = 3, bty = "n")
  
  # Summary Table (like precis)
  pred_summary <- data.frame(
    etape_length = etape_lengths,
    mean = apply(preds, 2, mean),
    sd   = apply(preds, 2, sd),
    `2.5%`  = apply(preds, 2, quantile, probs = 0.025),
    `97.5%` = apply(preds, 2, quantile, probs = 0.975)
  )
  
  print(pred_summary, row.names = FALSE)
  invisible(pred_summary) # Returns the summary silently for further use
}

```

```{r}
plot_depth_predictions(R_new = 500, post = post, etape_lengths = etape_lengths, colset = colset)
```

## etape effect model
```{r}
# ID effects model - does individual etape matter?
# Fit Model with Simulated Data
effect_model <- ulam(
  alist(
    # Observation model
    W ~ dnorm(mu, sigma_L),
    #W ~ dnorm(mu, sigma_L[L]),
    
    # Mean structure: interaction between analyte and sampler type
    mu <- alpha[L] + beta[L]*R + gamma[I],
    
    # Priors for eTape model-specific parameters
    alpha[L] ~ dnorm(aPrior, 1),  # Intercept for each etape length
    beta[L] ~ dnorm(-0.17, 1),  # Slope for each etape length
    gamma[I] ~ dnorm(0, 1),  # Individual etape effect
    # alpha[L] ~ dnorm(a_bar, sigma_a),
    # beta[L] ~ dnorm(b_bar, sigma_b),

    # Hyper-priors
    # a_bar ~ dnorm(45, 3),
    # sigma_a ~ dexp(1),
    # b_bar ~ dnorm(-0.0017, 1),      # Typical eTape: negative slope
    # sigma_b ~ dexp(1),
    
    # Prior for measurement error
    sigma_L ~ exponential(1)
    #vector[K_L]:sigma_L ~ exponential(1)
  ),
  data = d_real,
  chains = 4,
  cores = 4
)
```

### Effect Model Calibration Results - Real Data
```{r}
p_effect <- precis(effect_model, depth = 2, digits = 5)
p_effect
plot(p_effect)
```

```{r}
trankplot(effect_model)
```



## Frequentist Approach Results
```{r}
# Fit the model with interaction (intercept and slope by eTape length)
lm_fit <- lm(water_depth_cm ~ resistivity_ohm * etape_length, data = d)
# This is equivalent to: water_depth_cm ~ resistivity_ohm + etape_length + resistivity_ohm:etape_length

# View the summary
summary(lm_fit)
# confidence intervals for the coefficients
confint(lm_fit)
```

# Single eTape study
To help verify if the Bayesian coefficients are sufficiently accurate for practical use, a sub-study was performed where 4 student employees calibrated the same, 18-inch eTape. These new data will be used to create a linear model without considering the person effects.  If the uncertainty between the bayesian analysis over all etapes is similar to the uncertainty in the single eTape study, then it would suggest that the Bayesian model coefficients are accurate enough for practical use because they could not be statistically differentiated from the single eTape study.

```{r}
# Load the single eTape study data
single_etape_data <- read.csv("../data/single_etape_study.csv")

# clean data like previously
single_etape_data <- single_etape_data %>%
  mutate(
    water_depth_cm = water_depth_inch * 2.54,
    resistivity_ohm = as.numeric(resistivity_ohm),
    etape_ID = as.factor(etape_ID)
  )

# make data into list for ulam
d_single_etape <- list(
  W = single_etape_data$water_depth_cm,
  R = single_etape_data$resistivity_ohm,
  L = as.factor(single_etape_data$etape_ID),
  K_L = length(unique(single_etape_data$etape_ID)),
  aPrior = c(28, 32, 40, 48, 61)  # Prior for alpha (intercept) for each etape length
)

```

## Frequentist approach (a quick look)
```{r}
# fit a linear model to the single eTape study data
single_etape_lm <- lm(water_depth_cm ~ resistivity_ohm, data = single_etape_data)

# View the summary of the linear model
summary(single_etape_lm)

# confidence intervals for the coefficients
confint(single_etape_lm)
```

## Bayesian approach

```{r}
# Fit Model with Simulated Data
m2.0 <- ulam(
  alist(
    # Observation model
    W ~ dnorm(mu, sigma),
    
    # Mean structure: interaction between analyte and sampler type
    mu <- alpha + beta*R,
    
    # Priors for eTape model-specific parameters
    alpha ~ normal(50, 1),  # Intercept for each etape length
    beta ~ normal(-0.17, 1),  # Slope for each etape length
    
    # Prior for measurement error
    sigma ~ exponential(1)

  ),
  data = d_single_etape,
  chains = 4,
  cores = 4
)
```

### Single eTape Study Calibration Results
```{r}
#trankplot(m2.0)
#traceplot(m2.0)
single.effect <- precis(m2.0, depth = 2, digits = 5)
print(single.effect)
plot(single.effect)
```

```{r}
# compare precis output from real_model to m2.0

# compare at 95% confidence
p95 <- precis(real_model, prob = 0.95, depth = 2)
single.effect95 <- precis(m2.0, prob = 0.95, depth = 2)

comp_table <- data.frame(
  Parameter = rep(c("alpha", "beta"), each = 2),
  Model = c("All eTapes", "Single eTape", "All eTapes", "Single eTape"),
  Mean = c(
    p95$mean[4], single.effect95$mean[1],
    p95$mean[9], single.effect95$mean[2]
  ),
  lower = c(
    p95$`2.5%`[4], single.effect95$`2.5%`[1],
    p95$`2.5%`[9], single.effect95$`2.5%`[2]
  ),
  upper = c(
    p95$`97.5%`[4], single.effect95$`97.5%`[1],
    p95$`97.5%`[9], single.effect95$`97.5%`[2]
  )
)

print(comp_table)

alpha_table <- subset(comp_table, Parameter == "alpha")
beta_table  <- subset(comp_table, Parameter == "beta")
alpha_table$y <- 2:1
beta_table$y  <- 2:1

par(mar = c(5, 7, 4, 2) + 0.1)  # more space on the left

plot(
  x = alpha_table$Mean,
  y = alpha_table$y,
  xlim = range(c(alpha_table$lower, alpha_table$upper)),
  ylim = c(0.5, 2.5),
  xlab = "Alpha Estimate (mean and 95% CI)", ylab = "",
  yaxt = "n", pch = 16, cex = 1.5,
  main = "Comparison of alpha (intercept) Credible Intervals"
)
segments(
  x0 = alpha_table$lower,
  y0 = alpha_table$y,
  x1 = alpha_table$upper,
  y1 = alpha_table$y,
  lwd = 2
)
points(alpha_table$Mean, alpha_table$y, pch = 16, cex = 1.5)
axis(2, at = alpha_table$y, labels = alpha_table$Model, las = 1)
abline(h = alpha_table$y, col = "gray90", lty = 2)

plot(
  x = beta_table$Mean,
  y = beta_table$y,
  xlim = range(c(beta_table$lower, beta_table$upper)),
  ylim = c(0.5, 2.5),
  xlab = "Beta Estimate (mean and 95% CI)", ylab = "",
  yaxt = "n", pch = 16, cex = 1.5,
  main = "Comparison of beta (slope) Credible Intervals"
)
segments(
  x0 = beta_table$lower,
  y0 = beta_table$y,
  x1 = beta_table$upper,
  y1 = beta_table$y,
  lwd = 2
)
points(beta_table$Mean, beta_table$y, pch = 16, cex = 1.5)
axis(2, at = beta_table$y, labels = beta_table$Model, las = 1)
abline(h = beta_table$y, col = "gray90", lty = 2)


```

### compare results using generative model predictions
```{r}

# For real_model: extract alpha[4], beta[4]
alpha_real <- post$alpha[,4]
beta_real  <- post$beta[,4]

# For m2.0: extract alpha, beta (no index needed)
post_single <- extract.samples(m2.0)
alpha_single <- post_single$alpha
beta_single  <- post_single$beta

```

```{r}
compare_prediction_distributions <- function(
  R_new,
  post_real,   # extract.samples(real_model)
  post_single, # extract.samples(m2.0)
  etape_index_real = 4, # which alpha/beta in real_model to use,
  real_cm, # cm value for the vertical line
  col_real   = "#8080FF",
  col_single = "#F98400"
) {
  # Compute predictions
  preds_real   <- post_real$alpha[, etape_index_real] + post_real$beta[, etape_index_real] * R_new
  preds_single <- post_single$alpha + post_single$beta * R_new
  
  # Density objects
  dens_real   <- density(preds_real)
  dens_single <- density(preds_single)
  
  # Set up plot limits
  all_preds <- c(preds_real, preds_single)
  xlim <- range(all_preds)
  ylim <- c(0, max(dens_real$y, dens_single$y) * 1.1)
  
  # Plot
  plot(
    dens_real, type = "l", lwd = 3, col = col_real,
    xlim = xlim, ylim = ylim,
    xlab = "Predicted Water Depth (cm)",
    ylab = "Density",
    main = sprintf("Posterior Water Depth Distributions\nResistivity = %s 立\n(True value as Gray Line)", R_new)
  )
  lines(dens_single, lwd = 3, col = col_single, lty = 1)
  
  abline(v = real_cm, col = "gray", lty = 2, lwd = 3)  # vertical line at 0 for reference
  
  legend(
    "topright",
    legend = c("All eTapes", "Single eTape"),
    col = c(col_real, col_single),
    lwd = 3, lty = c(1, 1), bty = "n"
  )
  
  # Print summary table
  pred_summary <- data.frame(
    Model = c("All eTapes", "Single eTape"),
    mean = c(mean(preds_real), mean(preds_single)),
    sd   = c(sd(preds_real), sd(preds_single)),
    `2.5%` = c(quantile(preds_real, 0.025), quantile(preds_single, 0.025)),
    `97.5%` = c(quantile(preds_real, 0.975), quantile(preds_single, 0.975))
  )
  print(pred_summary, row.names = FALSE)
  invisible(pred_summary)
}

```


```{r}
compare_prediction_distributions(
  R_new = 1118,              # any resistivity value you want
  post_real = post,
  post_single = post_single,
  etape_index_real = 4,      # change if you want a different eTape group
  real_cm = 33.02            # cm value for the vertical line
)
```


```{r}
# Generate linear regression line plot
plot_both_regressions <- function(R_seq, post_real, post_single, etape_index_real = 4) {
  pred_mat_real <- sapply(R_seq, function(R) post_real$alpha[,etape_index_real] + post_real$beta[,etape_index_real] * R)
  pred_mean_real  <- apply(pred_mat_real, 2, mean)
  pred_lower_real <- apply(pred_mat_real, 2, quantile, 0.025)
  pred_upper_real <- apply(pred_mat_real, 2, quantile, 0.975)
  
  pred_mat_single <- sapply(R_seq, function(R) post_single$alpha + post_single$beta * R)
  pred_mean_single  <- apply(pred_mat_single, 2, mean)
  pred_lower_single <- apply(pred_mat_single, 2, quantile, 0.025)
  pred_upper_single <- apply(pred_mat_single, 2, quantile, 0.975)
  
  plot(
    R_seq, pred_mean_real, type = "l",
    col = "#8080FF", lwd = 2,
    ylim = range(c(pred_lower_real, pred_upper_real, pred_lower_single, pred_upper_single)),
    xlab = "Resistivity (立)",
    ylab = "Predicted Water Depth (cm)",
    main = "Regression with Posterior Uncertainty\nAll eTapes vs Single eTape"
  )
  polygon(
    c(R_seq, rev(R_seq)),
    c(pred_upper_real, rev(pred_lower_real)),
    col = adjustcolor("#8080FF", alpha.f = 0.2), border = NA
  )
  lines(R_seq, pred_mean_single, col = "#F98400", lwd = 2, lty = 2)
  polygon(
    c(R_seq, rev(R_seq)),
    c(pred_upper_single, rev(pred_lower_single)),
    col = adjustcolor("#F98400", alpha.f = 0.2), border = NA
  )
  lines(R_seq, pred_mean_real, col = "#8080FF", lwd = 2)
  lines(R_seq, pred_mean_single, col = "#F98400", lwd = 2, lty = 2)
  legend("topright",
         legend = c("All eTapes", "Single eTape"),
         col = c("#8080FF", "#F98400"),
         lwd = 2,
         lty = c(1,2),
         bty = "n",
         fill = adjustcolor(c("#8080FF", "#F98400"), alpha.f = 0.2)
  )
  range(pred_upper_real - pred_lower_real)
  range(pred_upper_single - pred_lower_single)
}




```

```{r}
R_seq <- seq(500, 2900, by = 100)
plot_both_regressions(R_seq, post, post_single)

```

